An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-09-21 report.tex
updates to report

@@

    A graphics processing unit (GPU) consists of several (thousand)
    compute cores and is designed for high-throughput computations.

    NVIDIA cards support extensions such as CUDA and OpenCL for
    langauages like C, C++, Fortran and Python.
    Applications written using these extensions are launched
    with thousands of lightweight threads,
    each operating on a small portion the problem domain (typically a single grid point).

    Each GPU has several so-called streaming multiprocessors (SMs)---the Tesla K20M
    has 13 SMs---each of which has its own control units, registers, caches, etc.
    Instructions execute simultaneously on an SM.

    All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
    Access to global memory is efficient when concurrent threads access neighbouring memory locations
    (known as \emph{coalesced memory access}).

    All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
    Access to shared memory, in general, is much faster than access to global memory,
    but the amount of shared memory per SM is limited, typically a few dozen KB.
    Certain memory access patterns by concurrent threads to shared memory may result in
    serialization of the memory accesses, known as bank conflicts. These are detailed [here].

    The OpenCL programming platform

@@

@@

    While evaluating the derivative, each block is assigned to a ``compute unit'',
    which may be a single core, a multi-core processor, or a GPU.
    Each block may be thought of as a collection of ``lines'' of grid points
    \ref{fig:one-line} in the x-direction,
    and the entire domain may be thought of as a collection of ``lines'' of
    blocks.
    While computing derivatives in other directions,
    (e.g., z-direction), we interpret each block as a collection of ``lines''
    of grid points in the z-direction and the domain as a collection of
    block in the z-direction.

    A GPU affords massive parallelism by virtue of several (thousand)
    lightweight compute threads (referred to as \emph{work items} in the OpenCL specification).
    The work items on a GPU are scheduled as \emph{blocks},
    which can be 1-D, 2-D, or 3-D.
    Each block is assigned to a \emph{streaming microprocessor}
    on the GPU---the Tesla K20 GPU has 13 streaming microprocessors---
    and runs in parallel.
    Threads within a block have common access to small, fast,
    \emph{shared memory}, which can be explicitly controlled.

@@

@@

\section{Results and performance comparison}

@@

    It is instructive also to examine the the total time

    We present the results for both

@@

% left, lower, right, upper
\begin{figure}[h]
\begin{center}
\includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/domain.eps}
\end{center}
\caption{Domain}
\label{fig:domain}
\end{figure}

\begin{figure}[ht]
\begin{minipage}[b]{0.45\linewidth}
\centering
\includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
\caption{Single block}
\label{fig:one-block}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}[b]{0.45\linewidth}
\centering
\includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
\caption{Single thread}
\label{fig:one-line}
\end{minipage}
\end{figure}

>>>>

    [FastTridiagonal] discuss the GPU implementation of three algorithms
    and their hybrid variants: Cyclic Reduction, Parallel Cyclic Reduction
    and Recursive Doubling.
    Their approach utilizes the GPUs \emph{shared memory},
    and addresses the concerns of bank conflicts arising in these algorithms.
    A global memory implementation of their solvers
    runs about three times slower, but is able to accomodate larger systems.
    The speedups they report are 12x over a 2.5 GHz quad-core CPU.

    [ScalableTridiagonal] and [Autotuned] both propose hybrid
    Parallel Cyclic Reduction and thread-level parallel Thomas algorithm (p-Thomas), and
    methods to determine the switching point for the algorithms.
    The speedups reported are 8.3x and 11x over a 3.33 GHz Intel i7 quad-core
    processor and a 3.4 GHz Intel i5 dual-core respectively.

    In [ScalableNumericallyStable],
    the authors implement a heterogenous (GPU+MPI) solver using the
    SPIKE algorithm for GPUs with a diagonal pivoting strategy for stability.
    More recently, [ChunkedCyclic] propose a \emph{chunked} cyclic reduction
    strategy for GPUss
    In [GPUApplicationForHigherOrderCFD], the inverse of the coefficient matrix
    is used to solve tridiagonal systems resulting from high-order compact
    finite difference schemes. The work here is for a single GPU.
    [An efficient GPU implementation of cyclic reduction solver] demonstrate
    a method to improve memory accesses for Cyclic Reduction
    using a reordering strategy, and apply it to compressible viscous flow
    simulations.

    --- Mention the speedups

    In this work, we propose a variant of Cyclic Reduction for solving
    tridiagonal systems arising in compact finite differences
    in a cluster environment (several GPU nodes).
    We introduce a strategy to
    reduce global memory accesses and decrease memory requirements by
    pre-computing coefficients appearing in Cyclic Reduction.

@@

    GPUs afford massive parallelism by virtue of several (thousand)
    lightweight compute threads, referred to as \emph{work items}.
    The work items on a GPU are scheduled as \emph{blocks},
    which can be 1-D, 2-D, or 3-D.
    Each block is assigned to a \emph{streaming multiprocessor} (SM)
    on the GPU---the Tesla K20 GPU has 13 streaming multiprocessor---and
    all threads in a block are executed simultaneously by the SM.
    All SMs (and therefore all threads) have access to the GPU's ``global'' memory.
    Access to global memory is efficient when concurrent threads
    access neighbouring memory locations
    (known as \emph{coalesced memory access}).
    All threads in an SM also have access to a common fast, on-chip, \emph{shared memory}.
    Access to shared memory, in general, is much faster than access to global memory,
    but the amount of shared memory per SM is limited, typically a few dozen KB.
    Certain memory access patterns by concurrent threads to shared memory may result in
    serialization of the memory accesses, known as bank conflicts. These are detailed [here].

@@

    % left, lower, right, upper
    \begin{figure}[h]
    \begin{center}
    \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/domain.eps}
    \end{center}
    \caption{Problem domain}
    \label{fig:domain}
    \end{figure}

    \begin{figure}[ht]
    \begin{minipage}[b]{0.45\linewidth}
    \centering
    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-block.eps}
    \caption{Single block assigned to a process}
    \label{fig:block}
    \end{minipage}

    \hspace{0.5cm}
    \begin{minipage}[b]{0.45\linewidth}
    \centering
    \includegraphics[trim={{50pt} {150pt} {50pt} {150pt}}, clip, height=200pt]{img/one-thread.eps}
    \caption{Single grid line within a block}
    \label{fig:grid-line}
    \end{minipage}
    \end{figure}

@@

    While evaluating the derivative in the x-direction (say),
    each block is assigned to a \emph{compute unit}, or a \emph{process}
    which may be a single core, a multi-core processor, or a GPU.
    Each block may be thought of as a collection of ``lines'' of grid points
    \ref{fig:one-line}---referred to as \emph{grid lines}---in the x-direction,
    and the entire domain may be thought of as a collection of such blocks.

@@

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{100pt} {150pt} {100pt} {150pt}}, clip, height=250pt]{img/block-dimensions.eps}
        \end{center}
        \caption{Domain}
        \label{fig:block-dimensions}
        \end{figure}

        In the reference approach, the domain is decomposed into several blocks,
        and each block is assigned to a CPU core.
        Each core operates on a block with dimensions as shown
        in the figure \ref{fig:block-dimensions}.
        The computational pattern followed by each core for the steps listed in Section[SEC]
        is a triple nested loop over the grid points in the block,
        followed by a communication of the ``faces'' of the block with the other
        cores:

        \begin{listing}
        \begin{minted}{fortran}
           do j=1,n2
              do i=1,n1
                 x(1,i,j) = bet(1)*x(1,i,j)
                 do k = 2, n3
                    x(k,i,j) = (x(k,i,j) - a(k)*x(k-1,i,j))*bet(k)
                 end do
                 p_loc(i,j) = x(n3,i,j)
              end do
           end do

           call MPI_Allgather(p_loc,n1*n2,double_real,p,n1*n2, &
              double_real,comm,error)

        \end{minted}
        \caption{Computational pattern in reference approach}
        \end{listing}

        Conceptually, the innermost loop solves the tridiagonal system
        for a single ``line'' of grid points.

        In a multi-threaded environment,
        rather than having a single thread perform the local part
        of the computations for each line,
        it is beneficial to have several threads ``sweep'' through the domain,
        each performing the computations for a different tridiagonal system.
        This is shown in Figure \ref{fig:block-multi-threaded} for a 4-core processor.
        In each iteration of the inner-most loop,
        4 lines of grid points are solved.

        \begin{figure}[h]
        \begin{center}
        \includegraphics[trim={{50pt} {250pt} {50pt} {150pt}}, clip, height=300pt]{img/block-multi-threaded.eps}
        \end{center}
        \caption{Domain}
        \label{fig:block-multi-threaded}
        \end{figure}

        Algorithms like LU and the Thomas Algorithm are inherently sequential.
        While thread level parallelism can be used to solve several systems at once,
        the number of steps to solve each system remains large.

        Algorithms like Cyclic Reduction and Parallel Cyclic Reduction use
        several parallel processors to solve a tridiagonal system.
        Cyclic reduction

\section{Description of the algorithm}

    First, we discuss the cluster level parallelism.

    Given N equations in N variables of the form:

    \begin{align}
    & A =
     \begin{bmatrix}
         b_1 & c_1 \\
         a_2 & b_2 & c_2 \\
             & a_3 & b_3 & c_3 \\
             &     & a_4 & b_4 & c_4 \\
             &     &     &     &  \ddots & c_{N-1}\\
             &     &     &     &     a_N  & b_N
      \end{bmatrix}&
    \end{align}

    each process $p$ is concerned with the solution of a \emph{subsystem}
    of equations:

    \begin{align}
    &L_p =
    \begin{bmatrix}
        b_1^p & c_1^p \\
        a_2^p & b_2^p & c_2^p \\
              & a_3^p & b_3^p & c_3^p \\
              &       & a_4^p & b_4^p & c_4^p \\
              &       &       &       &  \ddots & c_{M-1}^p\\
              &       &       &       &     a_{M}^p  & b_{M}^p
     \end{bmatrix}&
    \end{align}

    For each subsystem, the following equations are solved:

    \begin{align}
        & L_p\boldsymbol{x}_p^R = \boldsymbol{r_p} &  \label{eqn:local-eqn-1} \\
        & L_p\boldsymbol{x}_p^{UH} = (-a_1^p, 0, 0 \hdots 0)^T &  \label{eqn:local-eqn-2} \\
        & L_p\boldsymbol{x}_p^{LH} = (0, 0, 0, \hdots -c_M^p)^T \label{eqn:local-eqn-3} &
    \end{align}

    The general solution is obtained as a linear combination of the above solutions:

    \begin{equation} \label{eqn:linear-combo}
        \boldsymbol{x}_p = {\xi}_p^{UH} \boldsymbol{x}_p^R + {\xi}_p^{LH} \boldsymbol{x}_p^{LH}
    \end{equation}

    Where the parameters ${\xi}_p^{UH}$ and ${\xi}_p^{LH}$ are obtained
    by solving the following ``reduced'' system:

    \begin{equation} \label{eqn:reduced-system}
     \begin{bmatrix}
         x_{1,M}^{LH}   &   -1                                                          \\
         -1             &   x_{2,1}^{UH} &   x_{2,1}^{LH}                               \\
                        &   x_{2,M}^{LH} &   x_{2,M}^{LH}  &  -1                        \\
         &              &   -1           &   x_{3,M}^{UH}  &  x_{1,M}^{LH}              \\
         &              &   &                x_{3,M}^{UH}  &  x_{1,M}^{LH} &    -1      \\
         &              &   &            &   &             \ddots   \\
         &              &   &            &   &             &        \ddots \\
         &  &   &   &   &   &   & -1 x_{P,1}^{UH}
      \end{bmatrix}
    \begin{bmatrix}
        \xi_1^{LH} \\
        \xi_2^{UH} \\
        \xi_2^{LH} \\
        \xi_3^{UH} \\
        \xi_3^{LH} \\
        \vdots \\
        \vdots \\
        \xi_P^{UH} \\
     \end{bmatrix}
     =
     \begin{bmatrix}
         x_{1,M}^R \\
         x_{2,1}^R \\
         x_{2,M}^R \\
         x_{3,1}^R \\
         x_{3,M}^R \\
         \vdots \\
         \vdots \\
         x_P^R \\
      \end{bmatrix}
  \end{equation}

    The solution procedure is then:

    \begin{enumerate}
        \item Each process solves the three equations
        \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}.
        \item Each process communicates the parameters required to
        assemble the ``reduced'' system eqn:reduced-system
        \item The reduced system is solved for the coupling coefficients
        $\xi_i^{UH}$ and $_i^{LH}$, which are sent to their respective processes
        (if required)
        \item Each process solves \ref{eqn:linear-combo} for the local part
        of the solution to the global system.
    \end{enumerate}

    In the context of solving tridiagonal systems arising in compact finite differeces,
    we note the following:

    \begin{enumerate}
        \item Each process is involved in solving \emph{several} equations,
            one for each grid line.
        \item
        \item
    \end{enumerate}

    * Each process is involved in solving a number of equations, and not a single one
    * The left hand side for all the equations is the same,
        - LHS of \ref{eqn:local-eqn-1}-\ref{eqn:local-eqn-3}
        - RHS of \ref{eqn:local-eqn-2} and \ref{eqn:local-eqn-3} are the same
        - The above two equations need only be solved for a single grid line
    * Only equation \ref{eqn:local-eqn-3} is solved for every grid line,
    which amounts to solving the same equation for several different RHS

    Cyclic Reduction
    Precomputing coefficients in cyclic reduction
    OpenCL implementations
    Shared memory and associated issues
    Repeated kernel invocations
    Final kernel listings

    Overall solution strategy

@@

\section{Results and performance comparison}
    Notes on performance comparison
        - Fair comparisons
        - Memory transfers
        - Time spent on writing code
        - Compiler optimizations
    Problem size v/s solution time
    Breakup of time taken for various problem sizes and nprocs
    Results for CPU

@@

@@
